### 2019-07-06 methodology/bayesian-networks.tex
Methodology > novel contributions

\subsubsection{Approximations}
Several heuristical approaches have been developed to be able to find good network structures in an efficient manner.
Examples of these are:
\begin{itemize}
  \item Chow-Liu, that builds a tree approximation of the probability distribution
  \item Greedy hill-climbing, that adds/removes/flips an edge at a time
  \item optimal reinsertion, that iteratively calculates the optimal $Markov blanket$ (the subset of all nodes that are sufficient to determine the value of another subset) of an ever-smaller subset of nodes
\end{itemize}

>>>>

\subsubsection{Approximations}
Several heuristical approaches have been developed to be able to find good network structures in an efficient manner.
Examples of these are:
\begin{itemize}
  \item Chow-Liu, that builds a tree approximation of the probability distribution
  \item Greedy Hill-Climbing, that adds/removes/flips an edge at a time
  \item Optimal Reinsertion, that iteratively calculates the optimal $Markov blanket$ (the subset of all nodes that are sufficient to determine the value of another subset) of an ever-smaller subset of nodes
\end{itemize}

### 2019-07-06 methodology/instruments.tex
Methodology > novel contributions

\section{Methods} 
\subsection{Libraries}
\subsubsection{Pomegranate}
\textit{pomegranate} (\cite{pomegranate}) is an open-source probabilistic models package for python.
Its core philosophy is that every probabilistic model, from Hidden Markov to Bayesian Network, can be seen as a probability distribution and, as such, can be flexibly composed into hierarchical mixture models (\cite{Schreiber2017}).
The package implements:
\begin{itemize}
	\item Probability Distributions
	\item General Mixture Models
	\item Hidden Markov Models
	\item Bayes Classifiers and Na{\"i}ve Bayes
	\item Markov Chains
	\item \textbf{Bayesian Networks}
	\item Factor Graphs
\end{itemize}

@@

pomegranate currently only supports Discrete Bayesian Networks so the random variable of each node must have a categorical distribution.

@@

\subsection{Algorithms}
\subsubsection{d-separation}
A na{\"i}ve implementation according to the definition (presented in Subsec. \ref{subsec:bayesiannetworks}) to check for d-separation between node $X$ and $Y$ would have a complexity in the order of the number of trails between $X$ and $Y$; this leads to an exponential in the size of the graph running time.
Luckily, \cite{koller2007dseparation} present what is a linear time algorithm to solve the problem.

@@

>>>>

\section{Methods} 
\subsection{Libraries}\label{subsec:libraries}
\subsubsection{Pomegranate}
\textit{pomegranate} (\cite{pomegranate}) is an open-source probabilistic models package for python.
Its core philosophy is that every probabilistic model, from Hidden Markov to Bayesian Network, can be seen as a probability distribution and, as such, can be flexibly composed into hierarchical mixture models (\cite{Schreiber2017}).
The package implements:
\begin{itemize}
	\item Probability Distributions
	\item General Mixture Models
	\item Hidden Markov Models
	\item Bayes Classifiers and Na{\"i}ve Bayes
	\item Markov Chains
	\item \textbf{Bayesian Networks}
	\item Factor Graphs
\end{itemize}

@@

Pomegranate currently only supports Discrete Bayesian Networks so the random variable of each node must have a categorical distribution.

@@

\subsection{Algorithms}
\todo{quanto dettagli implementativi devo mettere?  perchè ho un sacco di funzioni di visualizzazione etc}

\subsubsection{Model construction}
The data was given in \texttt{.xlsx} format and was imported using \texttt{panda}'s \texttt{read\_excel} function that returned a \texttt{DataFrame} object.
The so imported data was then preprocessed by dropping unwanted records and binning the remaining ones as described in Tab. \ref{tab:datasetpreprocess} .
The actual BN representation is learned at runtime by calling the \texttt{from\_samples} method of \texttt{pomegranate}'s \texttt{BayesianNetwork} to solve the structure learning problem (defined in Subsec. \ref{subsec:bnstructurelearning}).

\subsubsection{d-separation}
A na{\"i}ve implementation according to the definition (presented in Subsec. \ref{subsec:bayesiannetworks}) to check for d-separation between node $X$ and $Y$ would have a complexity in the order of the number of trails between $X$ and $Y$; this leads to an exponential in the size of the graph running time.
Luckily, \cite{koller2007dseparation} present what is a linear time algorithm to solve the problem.

@@

\subsubsection{Other Machine Learning methods}

### 2019-07-06 methodology/novel-contributions.tex
Methodology > novel contributions

\begin{algorithm}[htp!]
	\caption{Exhaustive pseudo-MPE algorithm}
	\label{alg:pseudo-mpe-exhaustive}
	\begin{algorithmic}
		\STATE $evidence = $ user selected $(state,value)$ tuples
		\STATE $MPE_tree = $ root MPE Tree in $evidence$
		\WHILE{True} 
			\STATE $mpe\_states$ = \texttt{next\_most\_probable\_states()}
			\IF{$mpe\_states$ is not empty}
				\STATE $next\_state$ = head of $mpe\_states$ 
				\STATE propose $next\_state$ to user \COMMENT{the least entropic state}
				\IF{the user refuses $next\_state$}
					\FOR{$alternative\_state$ in $mpe\_states \smallsetminus next\_state$}
						\STATE propose $alternative\_state$ to user \COMMENT{the next least entropic states}
						\IF{the user accepts $alternative\_state$}
							\STATE call \texttt{generate\_alternative\_branch()} on $MPE\_tree$ 
							\STATE add $alternative\_state$ to $MPE\_tree$
							\STATE $evidence = evidence \cup alternative\_state$
						\ELSE
							\STATE continue
						\ENDIF
					\ENDFOR
				\ELSE
					\STATE add $next\_state$ to $MPE\_tree$
					\STATE $evidence = evidence \cup next\_state$
				\ENDIF
			\ELSE 
				\STATE return
			\ENDIF
		\ENDWHILE
	\end{algorithmic}
\end{algorithm}

\begin{algorithm}[htp!]
	\caption{Independencies pseudo-MPE algorithm}
	\label{alg:pseudo-mpe-independencies}
	\begin{algorithmic}
		\STATE $evidence = $ user selected $(state,value)$ tuples
		\STATE $MPE_tree = $ root MPE Tree in $evidence$
		\WHILE{True} 
			\STATE $separated = $ \texttt{evidence\_d\_separation()} \COMMENT{based on evidence of previous step}
			\STATE $mpe\_states$ = \texttt{next\_most\_probable\_states()}
			\STATE $mpe\_states = mpe\_states \smallsetminus separated$ 
			\IF{$mpe\_states$ is not empty}
				\STATE $next\_state$ = head of $mpe\_states$ 
				\STATE propose $next\_state$ to user \COMMENT{the least entropic state}
				\IF{the user refuses $next\_state$}
					\FOR{$alternative\_state$ in $mpe\_states \smallsetminus next\_state$}
						\STATE propose $alternative\_state$ to user \COMMENT{the next least entropic states}
						\IF{the user accepts $alternative\_state$}
							\STATE call \texttt{generate\_alternative\_branch()} on $MPE\_tree$ 
							\STATE add $alternative\_state$ to $MPE\_tree$
							\STATE $evidence = evidence \cup alternative\_state$
						\ELSE
							\STATE continue
						\ENDIF
					\ENDFOR
				\ELSE
					\STATE add $next\_state$ to $MPE\_tree$
					\STATE $evidence = evidence \cup next\_state$
				\ENDIF
			\ELSE 
				\STATE return
			\ENDIF
		\ENDWHILE
	\end{algorithmic}
\end{algorithm}

@@

\subsubsection{Alternative Explanation Branches}

\subsubsection{\enquote{Pseudo-MPE} from Random Evidence} \label{subsubsec: pseudo-mpe-random}

@@

\subsubsection{Natural Language Explanation}
\todo{da fare e magari pensare anche a visual explanations?}

\subsubsection{Pairwise Correlations}
An interesting addition is an algorithm to measure the interrelatedness between pairs of variables.

\subsubsection{Network Plotting} \label{subsubsec:plot-model}

>>>>

The underlying Bayesian Network that represents the data set is learned and queried through the  \texttt{pomegranate} (see Subsec. \ref{subsec:libraries}) API but the great majority of all the code is completely custom-written.
This was necessary because \texttt{pomegranate}, while having a powerful backend, was found to be severely lacking in the breadth and flexibility of its API.
Many basic operations, such as the calculation of a joint distribution, were not available so the only way was to implement lower-level workarounds while still using \texttt{pomegranate} for the most basic operations, as the calculation of a posterior distribution.
In particular, \texttt{dialogue} is implemented with the only direct calls to the API being when learning the network and when calling \texttt{predict\_proba}, that queries the \texttt{BayesianNetwork} object to calculate the posterior distribution of the states given the current evidence.
D-separation, in the second variant of the algorithm, is calculated via the \texttt{evidence\_d\_separation} procedure that implements the pseudocode presented in Alg. \ref{alg:d-separation}.

\begin{algorithm}[htp!]
	\caption{Exhaustive pseudo-MPE algorithm}
	\label{alg:pseudo-mpe-exhaustive}
	\begin{algorithmic}
		\STATE $evidence = $ user selected $(state,value)$ tuples
		\STATE $MPE\_tree = $ MPE Tree rooted in $evidence$
		\WHILE{True} 
			\STATE $mpe\_states$ = \texttt{next\_most\_probable\_states($evidence$)}
			\IF{$mpe\_states$ is not empty}
				\STATE $next\_state$ = head of $mpe\_states$ 
				\STATE propose $next\_state$ to user \COMMENT{the least entropic state}
				\IF{the user refuses $next\_state$}
					\FOR{$alternative\_state$ in $mpe\_states \smallsetminus next\_state$}
						\STATE propose $alternative\_state$ to user \COMMENT{the next least entropic states}
						\IF{the user accepts $alternative\_state$}
							\STATE call \texttt{generate\_alternative\_branch()} on $MPE\_tree$ 
							\STATE add $alternative\_state$ to $MPE\_tree$
							\STATE $evidence = evidence \cup alternative\_state$
						\ELSE
							\STATE continue
						\ENDIF
					\ENDFOR
				\ELSE
					\STATE add $next\_state$ to $MPE\_tree$
					\STATE $evidence = evidence \cup next\_state$
				\ENDIF
			\ELSE 
				\STATE return
			\ENDIF
		\ENDWHILE
	\end{algorithmic}
\end{algorithm}

\begin{algorithm}[htp!]
	\caption{Independencies pseudo-MPE algorithm}
	\label{alg:pseudo-mpe-independencies}
	\begin{algorithmic}
		\STATE $evidence = $ user selected $(state,value)$ tuples
		\STATE $MPE\_tree = $ MPE Tree rooted in $evidence$
		\WHILE{True} 
			\STATE $separated = $ \texttt{evidence\_d\_separation($evidence$)} \COMMENT{based on evidence of previous step}
			\STATE $mpe\_states$ = \texttt{next\_most\_probable\_states($evidence$)}
			\STATE $mpe\_states = mpe\_states \smallsetminus separated$ 
			\IF{$mpe\_states$ is not empty}
				\STATE $next\_state$ = head of $mpe\_states$ 
				\STATE propose $next\_state$ to user \COMMENT{the least entropic state}
				\IF{the user refuses $next\_state$}
					\FOR{$alternative\_state$ in $mpe\_states \smallsetminus next\_state$}
						\STATE propose $alternative\_state$ to user \COMMENT{the next least entropic states}
						\IF{the user accepts $alternative\_state$}
							\STATE call \texttt{generate\_alternative\_branch()} on $MPE\_tree$ 
							\STATE add $alternative\_state$ to $MPE\_tree$
							\STATE $evidence = evidence \cup alternative\_state$
						\ELSE
							\STATE continue \COMMENT{go to next proposal}
						\ENDIF
					\ENDFOR
				\ELSE
					\STATE add $next\_state$ to $MPE\_tree$
					\STATE $evidence = evidence \cup next\_state$
				\ENDIF
			\ELSE 
				\STATE return 
			\ENDIF
		\ENDWHILE
	\end{algorithmic}
\end{algorithm}

@@

\subsubsection{Alternative Explanation Branches}
The function to generate alternative branches to the main MPE branch in the dialogue tree is called after the user refuses a $(state,value)$ in the dialogue and accepts one of the the alternatives.
The motivation is to present the user with a simple \enquote{what-if} analysis, replying to the question \enquote{Had I accepted the $(state,value)$ presented me by the system, what would have been the most probable configuration of the remaining $(state,value)$ pairs?}.
The question is answered by generating a maximally probable, alternative MPE sub-branch rooted in the last node in the main MPE branch that is under construction through the dialogue.

The alternative branch is generated by what is essentially an automated version of \texttt{dialogue} that always accepts the first suggestion returned by \texttt{next\_most\_probable\_states}.
Given that \texttt{dialogue} and \texttt{generate\_alternative\_branch} are essentially one and the same, the latter inherits the same pruning strategies as the former.
That is, \texttt{generate\_alternative\_branch} called during the exhaustive dialogue will generate a maximally likely assignment over all variables in $V \smallsetminus E$ while when invoked from one of the other two variant of the dialogue algorithm, it will apply their same pruning strategies.

The implementation of the MPE Tree was based on the \texttt{Anytree} python package (see Subsec. \ref{subsec:libraries}).
This lightweight package lacks in flexibility but makes the creation of a tree very simple.
Creation of a new node is done by simply passing a pointer to the parent node to the \texttt{Node} object constructor.
Construction of trees composed of chains, as are the ones I am dealing with, can be done by simply keeping track of the pointer to the last inserted node.

An example of the output shown to the user at each step of the dialogue is shown in Fig. \ref{fig:alternative-branch_1} while the pseudocode for the three variants are shown in Alg. \ref{alg:alternative-branch-echaustive}, \ref{alg:alternative-branch-independencies} and \ref{alg:alternative-branch-thresholded}.
Note that $alternative\_evidence$ is local to this algorithm and is separate from the $evidence$ used in the main \texttt{dialogue} procedure.

\begin{figure}[htbp]
\centerline{\includegraphics[width=\columnwidth]{methodology/images/alternative-explanation-tree-example}}
\caption{Example output at the end of the d-separation-aware variant of \texttt{dialogue}.
	The tuple (\enquote{ki67},\enquote{<14}) was proposed but the expert refused it and accepted the alternative (\enquote{c erbB 2},\enquote{0}).
	The main MPE Branch has ID 1 while the \enquote{what-if} one has ID 2.}
\label{fig:alternative-branch}
\end{figure}

\begin{algorithm}[htp!]
	\caption{Exhaustive alternative explanation branch algorithm}
	\label{alg:alternative-branch-echaustive}
	\begin{algorithmic}
		\STATE $alternative\_evidence = evidence$ 
		\STATE $alt\_node = $ last node in the main MPE Tree
		\STATE $branch\_id = branch\_id + 1$
		\WHILE{True} 
			\STATE $mpe\_states$ = \texttt{next\_most\_probable\_states($alternative\_evidence$)}
			\IF{$mpe\_states$ is empty}
				\STATE return
			\ELSE
				\STATE $next\_state$ = head of $mpe\_states$ 
				\STATE create $next\_state$ node, tag it with $branch\_id$ and make it son of $alt\_node$
				\STATE update $alt\_node$ node to be $next\_state$ node
				\STATE $alternative\_evidence = alternative\_evidence \cup next\_state$
			\ENDIF
		\ENDWHILE
	\end{algorithmic}
\end{algorithm}

\begin{algorithm}[htp!]
	\caption{Independencies alternative explanation branch algorithm}
	\label{alg:alternative-branch-independencies}
	\begin{algorithmic}
		\STATE $alternative\_evidence = evidence$ 
		\STATE $alt\_node = $ last node in the main MPE Tree
		\STATE $branch\_id = branch\_id + 1$
		\WHILE{True} 
			\STATE $separated = $ \texttt{evidence\_d\_separation($alternative\_evidence$)}
			\STATE $mpe\_states$ = \texttt{next\_most\_probable\_states($alternative\_evidence$)}
			\STATE $mpe\_states = mpe\_states \smallsetminus separated$ 
			\IF{$mpe\_states$ is empty}
				\STATE return
			\ELSE
				\STATE $next\_state$ = head of $mpe\_states$ 
				\STATE create $next\_state$ node, tag it with $branch\_id$ and make it son of $alt\_node$
				\STATE update $alt\_node$ node to be $next\_state$ node
				\STATE $alternative\_evidence = alternative\_evidence \cup next\_state$
			\ENDIF
		\ENDWHILE
	\end{algorithmic}
\end{algorithm}

\todo{implementare e fare}
\begin{algorithm}[htp!]
	\caption{Independencies alternative explanation branch algorithm}
	\label{alg:alternative-branch-thresholded}
	\begin{algorithmic}
		\STATE $evidence = $ user selected $(state,value)$ tuples
		\STATE $MPE\_tree = $ root MPE Tree in $evidence$
		\WHILE{True} 
			\STATE $separated = $ \texttt{evidence\_d\_separation()} \COMMENT{based on evidence of previous step}
			\STATE $mpe\_states$ = \texttt{next\_most\_probable\_states()}
			\STATE $mpe\_states = mpe\_states \smallsetminus separated$ 
			\IF{$mpe\_states$ is not empty}
				\STATE $next\_state$ = head of $mpe\_states$ 
				\STATE propose $next\_state$ to user \COMMENT{the least entropic state}
				\IF{the user refuses $next\_state$}
					\FOR{$alternative\_state$ in $mpe\_states \smallsetminus next\_state$}
						\STATE propose $alternative\_state$ to user \COMMENT{the next least entropic states}
						\IF{the user accepts $alternative\_state$}
							\STATE call \texttt{generate\_alternative\_branch()} on $MPE\_tree$ 
							\STATE add $alternative\_state$ to $MPE\_tree$
							\STATE $evidence = evidence \cup alternative\_state$
						\ELSE
							\STATE continue \COMMENT{go to next proposal}
						\ENDIF
					\ENDFOR
				\ELSE
					\STATE add $next\_state$ to $MPE\_tree$
					\STATE $evidence = evidence \cup next\_state$
				\ENDIF
			\ELSE 
				\STATE return 
			\ENDIF
		\ENDWHILE
	\end{algorithmic}
\end{algorithm}

\todo{ancora da implementare true MPE}
\subsubsection{\enquote{Pseudo-MPE} from Random Evidence} \label{subsubsec: pseudo-mpe-random}
In order to compare the Pseudo-MPE output with the true MPE solution I implemented a simple algorithm that, starting from a random initial set of evidence, generates the relative pseudo-MPE and MPE solutions.
The random initial evidence set is constructed by randomly choosing a number in the interval $k = [ 1, |V| ]$, with $V$ the set of vertices in the BN, and then randomly selecting $k$ of the random variables in $V$ to yield the set of variables $E$.
The value of each variable is randomly chosen among the set of its values; as all variables are categorical, this can easily be done.

\todo{vedere perchè magari sposto tutta l'implementazione su MPE\_Graph quando metto evidenze multiple ovunque}
The implementation is based on the \texttt{NetworkX} python library (see Subsec. \ref{subsec:libraries}) package as what is being constructed is not a tree but a \textit{polytree} (see Subsec. \ref{subsec:graph-theory}), as nodes may have multiple parents.
\texttt{NetworkX} is a much more complete package that \texttt{Anytree} and this allows greater flexibility, at the cost of slightly more complexity.
Note that $alternative\_evidence$ is considered separate from the main $evidence$ used in \texttt{dialogue}.

The pseudocode is shown in Alg. \ref{alg:pseudo-mpe-random-evidence} while an example output can be seen in Fig. \ref{fig:pseudo-mpe-random}.

\begin{algorithm}[htp!]
	\caption{Pseudo-MPE from random evidence algorithm}
	\label{alg:pseudo-mpe-random-evidence}
	\begin{algorithmic}
		\STATE $evidence = \emptyset$
		\STATE generate random number $k \in [ 1, |V| ]$
		\STATE $S = $ choose $k$ variables from $V$
		\FOR{$s \in S$}
			\STATE choose random $v$ in the possible values of $e$ 
			\STATE $evidence = evidence \cup (s,v)$
		\ENDFOR
		\STATE $MPE\_tree = $ MPE Tree rooted in $evidence$
		\STATE $last\_node = evidence$
		\STATE $alternative\_evidence = evidence$ 
		\WHILE{True} 
			\STATE $mpe\_states$ = \texttt{next\_most\_probable\_states($alternative\_evidence$)}
			\IF{$mpe\_states$ is empty}
				\STATE return
			\ELSE
				\STATE $next\_state$ = head of $mpe\_states$ 
				\STATE create $next\_state$ node and make it son of $last\_node$
				\STATE update $alt\_node$ node to be $next\_state$ node
				\STATE $alternative\_evidence = alternative\_evidence \cup next\_state$
			\ENDIF
		\ENDWHILE
	\end{algorithmic}
\end{algorithm}

\begin{figure}[htbp]
\centerline{\includegraphics[width=\columnwidth]{methodology/images/pseudo-mpe-random-example}}
\caption{Example output of the Pseudo-MPE from random evidence algorithm.}
\label{fig:pseudo-mpe-random}
\end{figure}

@@

\subsubsection{Natural Language Explanation}
\todo{da fare e magari pensare anche a visual explanations?}
The probability of each proposed tuple is quantified in natural language based on the probability of the most probable value within the variable.
These are shown in Tab. \ref{tab:naturallanguageprobabilities}.

\begin{table*}[htbp]
\caption{Probability quantifiers in natural language}
\begin{tabularx}{\textwidth}{@{} X X @{}}
\toprule 
Probability range & Natural language quantifier \\
\midrule 
(0, 0.2) &  "highly unlikely" \\
(0.2, 0.3) & "very unlikely" \\
(0.3, 0.4) & "unlikely" \\
(0.4, 0.5) & "not plausibly" \\
(0.5, 0.6) & "plausibly" \\
(0.6, 0.7) & "possibly" \\
(0.7, 0.8) & "likely" \\
(0.8, 0.9) & "very likely" \\
(0.9, 1) &  "highly likely" \\
(1) &  "certain" \\
\bottomrule
\end{tabularx}
\label{tab:naturallanguageprobabilities}
\end{table*}

\subsubsection{Pairwise Correlations} \todo{per ora non funziona}
An interesting addition is an algorithm to measure the interrelatedness between pairs of variables.
